P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2025-04-30 sage_latex_template/manuscript.tex
Update text for the compilation process
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

\DeclareRobustCommand{\RMchange}[1]{{\sethlcolor{cyan}\hl{#1}}}
%\newcommand{\RMchange}[1]{{\colorbox{purple!70}{#1}}}
%\newcommand{\RMchange}[1]{{\setlength{\fboxsep}{0pt}\colorbox{purple!70}{#1}}}
%\newcommand{\RMchange}[1]{#1}

@@

\autoref{fig:features} displays an overview of P4IRS.
In the left side, the front-end library from P4IRS is displayed with the available features, all these are defined through a Python interface by the user for modeling the simulation and determining the optimization schedule.
In the middle, the backend with the internal methods (deep-embedded in Python) with the compiler transformations and analyses are shown, these make use of the front-end representation to produce the final backend code.
In the right side, the final generated code is integrated and compiled together with the runtime modules and functionalities to achieve parallelism and accesses to external features.
The next subsections explore the different P4IRS features in more detail.

@@

Generating the final backend code for the specified kernels with proper parallelism, optimization strategies and GPU offloading is challenging.
P4IRS uses its own \ac{IR} to facilitate this process, where internal routines to build the cell lists, neighbor lists and generate the communication buffers are deep-embedded into Python to further be mapped into the P4IRS \ac{AST}.
Afterwards, compiler analyses and transformations are performed to properly add parallelism and optimizations into the code.
In its current state, P4IRS generates C++ and CUDA code with MPI for distributed-memory parallelism, but new backend languages, intermediate representations and technologies can be included without difficulties by extending its code generator.

@@

\autoref{lst:build_neigh} displays an example for constructing the intermediate representation that generates the code that builds the neighbor-lists for each particle in the system.
The \emph{BuildNeighborLists} class is a separate, high-level element from the AST in P4IRS.
During the compilation process, P4IRS performs analyses and transformations to lower this element by using its \emph{lower} method, and decorators such as the \emph{@pairs\_device\_block} are used to determine the target hardware where the code must be executed.

Different routines in P4IRS can be separated by ``modules'', and every module determines a point where data must be synchronized for the target it will execute on.
For example, the code generated from the constructed IR in the \emph{lower} method from \emph{BuildNeighborLists} is preceded by the appropriate copy calls into the GPU for all the arrays and properties that are referred within the IR subtree.
To avoid complex control-flow analyses to determine which properties and arrays actually have to be copied (since some of them can already be synchronized with the device), we provide bitmasks to represent each array and property defined in the simulation, and provide methods to check every array and property used within the intermediate representation.
Only when the bit is not set in the device, the copy method for the respective array or property is called.
Also, when properties or arrays are changed within the subtree, specific code is generated to clean up the respective bits.

The \emph{check\_resize} method is used to specify which array sizes can grow during the module execution, and which variable or array is used as their current capacity.
In this case, the \emph{neighbor\_capacity} can be increased based on the \emph{numneighs} array, so when one particle contains more neighbors than \emph{neighbor\_capacity}, a proper code is generated to reallocate the arrays that contain the referred capacity as one of its dimension sizes.
The compiler handles the reallocation cases when the target is an accelerator such as a GPU, which needs the kernel to be re-executed after arrays are reallocated since it is not possible to reallocate them directly in the accelerator.
P4IRS also takes care of avoiding memory access violation in the current kernel execution by wrapping the accesses around branches which assure the accessed indexes are smaller than the current capacity of the arrays.

Other elements such as the \emph{ParticleFor} and \emph{ParticleInteraction} are used as higher-level elements of the AST.
These will be lowered at some point based on their strategies and given parameters.
In this case, the \emph{ParticleInteraction} is lowered by using the linked-cells data structure to build the neighbor-lists as is done in traditional implementations.
The interaction object also contains methods to access the current particle (i), the neighbor particle (j) and the shape for the neighbor particle (shape).
The latter is particulary important for the shape-partitioning and multi-kernel code generation discussed in the \nameref{sec:shape_partitioning} section.

@@

After the IR is constructed for a specific routine, P4IRS then focus on determining which loops will be scheduled to the GPU.
As a general case, every outer-most loop in a block that uses the \emph{@pairs\_device\_block} decorator is a candidate to be replaced by a kernel launch to be run in parallel.
In \autoref{lst:build_neigh}, both loops at lines 9 and 13 are candidates and their body instructions are then wrapped around a separate GPU kernel.
The loops are then replaced by a kernel launch statement for the new generated kernel.
Note that some candidates may not be offloaded to the GPU, the reason is that P4IRS can use internal heuristics to determine whether a loop should or should not be executed in the GPU.
Domain-knowledge can also be used in the GPU offloading analyses since elements like \emph{ParticleFor} and \emph{ParticleInteraction} can be used during the compiler passes before they are lowered.

@@

\subsection{Runtime Module}

@@

To handle this, P4IRS provides a runtime module which contains such functionalities directly implemented in C++, and these make use of and synchronize (when necessary) the data structures from the generated simulation.
To access the simulation data, an instance (referred to as \emph{PairsRuntime} object) contains all the meta-data for the simulation, and can be used in these implemented routines.

@@

>>>>

%\DeclareRobustCommand{\RMchange}[1]{{\sethlcolor{cyan}\hl{#1}}}
%\newcommand{\RMchange}[1]{{\colorbox{purple!70}{#1}}}
%\newcommand{\RMchange}[1]{{\setlength{\fboxsep}{0pt}\colorbox{purple!70}{#1}}}
\newcommand{\RMchange}[1]{#1}

@@

\autoref{fig:features} displays an overview of P4IRS.
In the left side, the front-end library from P4IRS is displayed with the available features, all these are defined through a Python interface by the user for modeling the simulation and determining the optimization schedule.
In the middle, the backend with the internal methods (deep-embedded in Python) with the compiler transformations and analyses are shown, these make use of the front-end representation to produce the final backend code.
In the right side, the final generated code is integrated and compiled together with the runtime library and functionalities to achieve parallelism and accesses to external features.
The next subsections explore the different P4IRS features in more detail.

@@

\subsection{Intermediate Representation}

\RMchange{An important aspect of DSLs is to be able to use domain knowledge for their compiler transformations, thus enabling code optimizations that may not be detected by general purpose compilers.
In order to achieve this, P4IRS provides its own IR which can represent elements that are specific to particle computations.
These particle-specific elements in the AST can then be used in P4IRS compilation passes, and at some point during the compilation are lowered, resulting in the low level AST that corresponds to the code implementation that matches the strategies given by the user.}

@@

\RMchange{The P4IRS IR can be constructed in two ways: (a) by mapping of the Python IR generated by triggering the \emph{compile()} method and (b) by explicitly constructing the IR elements in-place.
The former is used for user-defined particle interactions (like the \emph{lennard\_jones} method from \autoref{lst:pairslj}) and the latter is normally used internally by P4IRS to construct the IR for internal routines.}
\autoref{lst:build_neigh} displays an example for constructing the intermediate representation that generates the code that builds the neighbor-lists for each particle in the system.
The \emph{BuildNeighborLists} class is a separate, high-level element from the AST in P4IRS.
During the compilation process, P4IRS performs analyses and transformations to lower this element by using its \emph{lower} method, and decorators such as the \emph{@pairs\_device\_block} are used to determine the target hardware where the code must be executed.

\RMchange{Note the domain-specific elements in the IR such as \emph{ParticleInteraction} and \emph{ParticleFor}, as already discussed these can be used during compilation passes for obtaining domain knowledge, and at some point they are lowered according to the strategies specified by the user.
For instance, the \emph{ParticleInteraction} abstract element can be lowered to a code representation that traverses pair of particles by either using the Linked Cells or Verlet Lists data structures.}
%Other elements such as the \emph{ParticleFor} and \emph{ParticleInteraction} are used as higher-level elements of the AST.
%These will be lowered at some point based on their strategies and given parameters.
In this case, the \emph{ParticleInteraction} is lowered by using the linked-cells data structure to build the neighbor-lists as is done in traditional implementations.
The interaction object also contains methods to access the current particle (i), the neighbor particle (j) and the shape for the neighbor particle (shape).
The latter is particulary important for the shape-partitioning and multi-kernel code generation discussed in the \nameref{sec:shape_partitioning} section.

@@

The final step of the P4IRS compiler is the code generation, which converts the optimized \ac{IR} to low-level hardware assembly instructions or to another high-level language (source-to-source compilation).
%P4IRS uses its own \ac{IR} to facilitate this process, where internal routines to build the cell lists, neighbor lists and generate the communication buffers are deep-embedded into Python to further be mapped into the P4IRS \ac{AST}.
%Afterwards, compiler analyses and transformations are performed to properly add parallelism and optimizations into the code.
In its current state, P4IRS generates C++ and CUDA code with MPI for distributed-memory parallelism, but new backend languages, intermediate representations and technologies can be included without difficulties by extending the code generator.
Although most of the code generation is straightforward for the \ac{IR} in its lowest level, there are some important points that are important to handle different backends and achieve performance-portability.

\RMchange{P4IRS creates ``modules'' for one or more routines when building its IR}, and every module determines a point where data must be synchronized for the target it will execute on.
For example, the code generated from the constructed IR in the \emph{lower} method from \emph{BuildNeighborLists} is preceded by the appropriate copy calls into the GPU for all the arrays and properties that are referred within the IR subtree.
To avoid complex control-flow analyses to determine which properties and arrays actually have to be copied (since some of them can already be synchronized with the device), we provide bitmasks to represent each array and property defined in the simulation, and provide methods to check every array and property used within the intermediate representation.
Only when the bit is not set in the device, the copy method for the respective array or property is called.
Also, when properties or arrays are changed within the subtree, specific code is generated to clean up the respective bits.

\RMchange{In addition,} the \emph{check\_resize} (see line 7 at \autoref{lst:build_neigh}) method is used to specify which array sizes can grow during the module execution, and which variable or array is used as their current capacity.
In this case, the \emph{neighbor\_capacity} can be increased based on the \emph{numneighs} array, so when one particle contains more neighbors than \emph{neighbor\_capacity}, a proper code is generated to reallocate the arrays that contain the referred capacity as one of its dimension sizes.
The compiler handles the reallocation cases when the target is an accelerator such as a GPU, which needs the kernel to be re-executed after arrays are reallocated since it is not possible to reallocate them directly in the accelerator.
P4IRS also takes care of avoiding memory access violation in the current kernel execution by wrapping the accesses around branches which assure the accessed indexes are smaller than the current capacity of the arrays.

\RMchange{When} the IR is constructed for a specific routine, P4IRS then focus on determining which loops will be scheduled to the GPU.
As a general case, every outer-most loop in a block that uses the \emph{@pairs\_device\_block} decorator is a candidate to be replaced by a kernel launch to be run in parallel.
In \autoref{lst:build_neigh}, both loops at lines 9 and 13 are candidates and their body instructions are then wrapped around a separate GPU kernel.
The loops are then replaced by a kernel launch statement for the new generated kernel.
Note that some candidates may not be offloaded to the GPU, the reason is that P4IRS can use internal heuristics to determine whether a loop should or should not be executed in the GPU.
Domain-knowledge can also be used in the GPU offloading analyses since elements like \emph{ParticleFor} and \emph{ParticleInteraction} can be used during the compiler passes before they are lowered.

@@

\subsection{Runtime Library}

@@

To handle this, P4IRS provides a runtime library which contains such functionalities directly implemented in C++, and these make use of and synchronize (when necessary) the data structures from the generated simulation.
To access the simulation data, an instance (referred to as \emph{PairsRuntime} object) contains all the meta-data for the simulation, and can be used in these implemented routines.

@@

\RMchange{The runtime library must be linked with the generated binary during its execution, and therefore it must be properly built for the target hardware and platform, and either installed in the target system or its path included in the list of paths where dynamic libraries are looked for (for example, the \emph{LD\_LIBRARY\_PATH} environment variable on Linux systems).
Although its compilation is done separately from the generated code and it should ideally be dynamically linked during runtime, it is possible to compile both together and statically link the runtime library code with the simulation code, hence generating a single, platform-specific, stand-alone binary that does not require further linking with the runtime code.}
